Carryover effects in free recall reveal how prior experiences influence memories of new experiences

### 2023-01-10 paper/main.tex
formatting tweaks...
finished a complete draft of the methods section   - note: the PDF isn't compiling correctly within VS Code...i'm not sure why...

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress, numbers, super]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}

@@

\caption{\textbf{Generating stochastic feature-sorted lists.} For a given
feature dimension (e.g., color), we compute the similarity
(Eqn.~\ref{eqn:similarity}) between the feature value(s) of the previous item,
$x$, and all yet-to-be-presented items ($a$ -- $g$). Next, we normalize these
similarity scores so that they sum to one. We lay in sequence a set of
``sticks,'' one for each candidate item, whose lengths are equal to these
normalized similarity scores. Note that the combined lengths of these sticks is
one. To select the next to-be-presented item, we draw a random number, $r$,
from the uniform distibution bounded between 0 and 1 (inclusive). The identity
of the next item is given by the stick adjacent to an indicator that moves
distance $r$ (starting from 0) along the sequence of sticks. In this case, the
next to-be-presented item is $e$. Note that each item's chances of selection is
proportional to its similarity to the previous item, along the given feature
dimension.} \label{fig:stick} 
\end{figure}

@@

@@

Following our prior work~\citep{HeusEtal17}, we used a permutation-correction
procedure to help isolate the behavioral aspects of clustering that we were
most interested in. After computing the uncorrected clustering score (for the
given list and observed recall sequence), we compute a ``null'' distribution of
$n$ additional clustering scores after randomly shuffling the recall order (we
use $n = 500$ in the present study). This null distribution represents an
approximation of the range of clustering scores one might expect to observe by
``chance,'' given that a hypothetical participant was \textit{not} truly
clustering their recalls, but where the hypothetical participant studied and
recalled exactly the same items (with the same features) as the true
participant.  We define the permutation-corrected clustering score as the percentile rank of the
observed uncorrected clustering score in this estimated null distribution.  In this way,
a corrected score of 1 indicates that the observed score was greater than any clustering score
one might expect by chance; in other words, good evidence that the participant was truly clustering their
recalls along the given feature dimension.

@@

@@

\paragraph{Ordering ``stabilize'' lists by an estimated fingerprint.}

\paragraph{Ordering ``destabilize'' lists by an estimated fingerprint.}

@@

\paragraph*{Temporal clustering score.}

Temporal clustering describes a participant's tendency to organize their recall
sequences by the learned items' encoding positions. For instance, if a
participant recalled the episode events in the exact order they occurred (or in
exact reverse order), this would yield a score of 1. If a participant recalled
the events in random order, this would yield an expected score of 0.5. For each
recall-event transition (and separately for each participant), we sorted all
not-yet-recalled events according to their absolute lag (that is, distance away
in the episode). We then computed the percentile rank of the next event the
participant recalled. We took an average of these percentile ranks across all
of the participant’s recalls to obtain a single (uncorrected) temporal clustering score for
the participant.  Finally, as with the feature-based clustering scores, we applied
our permutation correction procedure to obtain a corrected clustering score.

@@

@@

    \caption{\textbf{Memory ``fingerprints'' (order manipulation conditions).}
    The across-participant distributions of clustering scores for each feature
    type ($x$-coordinate) are displayed for each experimental condition
    (color), separately for order manipulation (early, top) and randomly
    ordered (late, bottom) lists. See
    Figures~\fingerprintsRandom~and~\fingerprintsAdaptive~for analogous plots
    for the random (control) and adaptive conditions.}
        \label{fig:fingerprints}
    \end{figure}

@@

    \caption{\textbf{Feature clustering carryover effects.} \textbf{A.} Feature clustering scores for ordder manipulation (early) versus randomly ordered (late) lists.
    \textbf{B.}  Accuracy differences (on early versus late lists) versus feature clustering ``carryover'' (defined as the differences between the average clustering scores on early and late lists).
    \textbf{C.}  Temporal clustering differences (on early versus late lists) versus feature clustering carryover.  \textbf{All panels.} Each dot in the main scatterplots denotes the
    average scores for one participant. The colored regression lines are computed
    across participants. The inset displays condition-averaged results, where each
    dot reflects a single condition and the regression line is computed across
    experimental conditions. All error ribbons denote bootstrap-estimated 95\%
    confidence intervals.} \label{fig:clustering-carryover}
    \end{figure}

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress, numbers, super]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{amsopn}

@@

\caption{\textbf{Generating stochastic feature-sorted lists.} For a given
feature dimension (e.g., color), we compute the similarity
(Eqn.~\ref{eqn:similarity}) between the feature value(s) of the previous item,
$x$, and all yet-to-be-presented items ($a$ -- $g$). Next, we normalize these
similarity scores so that they sum to one. We lay in sequence a set of
``sticks,'' one for each candidate item, whose lengths are equal to these
normalized similarity scores. Note that the combined lengths of these sticks is
one. To select the next to-be-presented item, we draw a random number, $r$,
from the uniform distibution bounded between 0 and 1 (inclusive). The identity
of the next item is given by the stick adjacent to an indicator that moves
distance $r$ (starting from 0) along the sequence of sticks. In this case, the
next to-be-presented item is $e$. Note that each item's chances of selection is
proportional to its similarity to the previous item, along the given feature
dimension.} \label{fig:stick}

@@

\paragraph*{Temporal clustering score (uncorrected).}

Temporal clustering describes a participant's tendency to organize their recall
sequences by the learned items' encoding positions. For instance, if a
participant recalled the episode events in the exact order they occurred (or in
exact reverse order), this would yield a score of 1. If a participant recalled
the events in random order, this would yield an expected score of 0.5. For each
recall-event transition (and separately for each participant), we sorted all
not-yet-recalled events according to their absolute lag (that is, distance away
in the episode). We then computed the percentile rank of the next event the
participant recalled. We took an average of these percentile ranks across all
of the participant’s recalls to obtain a single (uncorrected) temporal
clustering score for the participant.

@@

Following our prior work~\citep{HeusEtal17}, we used a permutation-correction
procedure to help isolate the behavioral aspects of clustering that we were
most interested in. After computing the uncorrected clustering score (for the
given list and observed recall sequence), we compute a ``null'' distribution of
$n$ additional clustering scores after randomly shuffling the recall order (we
use $n = 500$ in the present study). This null distribution represents an
approximation of the range of clustering scores one might expect to observe by
``chance,'' given that a hypothetical participant was \textit{not} truly
clustering their recalls, but where the hypothetical participant studied and
recalled exactly the same items (with the same features) as the true
participant. We define the permutation-corrected clustering score as the
percentile rank of the observed uncorrected clustering score in this estimated
null distribution. In this way, a corrected score of 1 indicates that the
observed score was greater than any clustering score one might expect by
chance; in other words, good evidence that the participant was truly clustering
their recalls along the given feature dimension.  We applied this correction procedure
to all of the clustering scores (feature and temporal) reported in this paper.

@@

We define each participant's \textit{memory fingerprint} as the set of their
permutation-corrected clustering scores across all dimensions we tracked in our
study, including their six feature-based clustering scores (category, size,
length, first letter, color, and location) and their temporal clustering score.
Conceptually, this memory fingerprint describes the participant's tendancies to
order (and, presumably, organize in memory) the studied words along each
dimension. To obtain stable estimates of these fingerprints for each
participant, we averaged clustering scores across lists. We also tracked and
characterized how participants' fingerprints changed across lists (e.g.,
Figs.~\ref{fig:fingerprint-trajectories},~\fingerprintTrajectoryRandom).

@@

The presentation orders of some lists in the adaptive condition of our
experiment (see \textit{Adaptive condition}) were sorted according to
participants' \textit{current} memory fingerprint, estimated using all of the
lists they had studied up to that point in the experiment. Because our
experiment incorporated a speech-to-text component, all of the behavioral data
for each participant could be analyzed just a few seconds after the conclusion
of the recall intervals for each list. We used the \texttt{Quail} Python
package~\citep{HeusEtal17} to apply speech-to-text alorithms to the just
collected data, aggregate the data for the given participant, and estimate the
participant's memory fingerprint using all of their available data up to that
point in the experiment. Two aspects of our implementation are worth noting.
First, because memory fingerprints are averaged across lists, the
already-computed memory fingerprints for earliar lists could be cached and
loaded as needed in future computations. This meant that our computations
pertaining to updating our estimates of a participant's memory fingerprint only
needed to consider data from the most recent list. Second, each element of the
null distributions of uncorrected fingerprint scores (see
\textit{Permutation-corrected feature clustering scores}) could be estimated
independently from the others. This enabled us to make use of the testing
computers' multi-core CPU archetectuers by elements of the null distributions
in batches of eight (i.e., the number of CPU cores on each testing computer).
Taken together, we were able to compress the fingerprint computations into just
a few seconds of computing time. The combined processing time for the
speech-to-text algorithm and fingerprint computations easily fit within the
inter-list intervals, where participants typically paused before moving on to
the next list.

\paragraph{Ordering ``stabilize'' and ``destabilize'' lists by an estimated
fingerprint.}

In the adaptive condition of our experiment, the presentation orders for
\textit{stabilize} and \textit{destabilize} lists were chosen to either
maximally or minimally (respectively) comport with participants' memory
fingerprints. Given a participant's memory fingerprint and a to-be-presented
set of items, we designed a permutation-based procedure for ordering the items.
First, we dropped from the participant's fingerprint the temporal clustering
score. For the remaining feature dimensions, we arranged the clustering scores
in the fingerprint into a template vector, $f$. Second, we computed $n = 2500$
random permutations of the to-be-presented items. These permutations served as
prospective presentation orders. We sought to select the specific order that
most (or least) matched $f$. Third, for each random permutation, we computed
the (permutation-corrected) ``fingerprint,'' treating the permutation as though
it were a potential ``perfect'' recall sequence. (We did not include temporal
clustering scores in these fingerprints.) This yielded a ``simulated fingerprint''
vector, $\hat{f}_p$ for each permutation $p$.  We used these simulated fingerprints to
select a specific permutation, $i$, that either maximized (for stabilized lists) or minimized (for destabilize lists)
the correlation between $\hat{f}_i$ and $f$.

Following some of our prior work~\citep{HeusEtal18a}, we use low-dimensional
embeddings to help visualize how participants' memory fingerprints change
across lists
(Figs.~\ref{fig:fingerprint-trajectories}A,~\fingerprintTrajectoryRandom A). To
compute a shared embedding space across participants and experimental
conditions, we concatenated the full set of fingerprints (across all lists,
participants, and experimental conditions) to create a large matrix with
number-of-lists $\times$ number-of-participants rows and seven columns (one for
each word feature dimension's clustering scores, plus an additional temporal
clustering score column). We used principal components analysis to project the
seven-dimensional observations into a two-dimensional space (using the two
principal components that explained the most variance in the data). For two
visualizations
(Figs.~\ref{fig:fingerprint-trajectories}B,~and~\fingerprintTrajectoryRandom B)
we computed an additional set of two-dimensional embeddings for participants'
\textit{average} fingerprints (i.e., across lists within a given group of
lists-- early or late). For those visualizations we averaged each participant's
rows (for the given group of lists) in the combined fingerprint matrix prior to
projecting it into the shared two-dimensional space. This yielded a single
two-dimensional coordinate for each \textit{participant} and \textit{list
group}, rather than for each individual list. We used these embeddings solely
for visualization. All statistical tests were carried out in the original
(seven-dimensional) feature spaces.

@@

@@

We used the distances between feature values for successively presented words
(see \textit{Defining feature-based distances}) to estimate ``event
boundaries'' where the feature values changed more than
usual~\citep{EzzyDava11, MannEtal16,RadvCope06, SwalEtal09, SwalEtal11,
DuBrDava16}. For each list, for each feature dimension, we computed the
distribution of distances between the feature values for successively presented
words. We defined event boundaries (e.g., Fig.~\ref{fig:recall-dynamics}B) as
occuring between any successive pair of words whose distances along the given
feature dimension were greater than one standard deviation above the mean for
that list. Note that, because event boundaries are defined for each feature
dimension, each individual list may contain several sets of event boundaries,
each at different moments in the presentation sequence (depending on the
feature dimension of interest).

@@

\caption{\textbf{Memory ``fingerprints'' (order manipulation conditions).} The
across-participant distributions of clustering scores for each feature type
($x$-coordinate) are displayed for each experimental condition (color),
separately for order manipulation (early, top) and randomly ordered (late,
bottom) lists. See Figures~\fingerprintsRandom~and~\fingerprintsAdaptive~for
analogous plots for the random (control) and adaptive conditions.}
\label{fig:fingerprints}

@@

    \caption{\textbf{Feature clustering carryover effects.} \textbf{A.} Feature
    clustering scores for ordder manipulation (early) versus randomly ordered
    (late) lists. \textbf{B.} Accuracy differences (on early versus late lists)
    versus feature clustering ``carryover'' (defined as the differences between
    the average clustering scores on early and late lists). \textbf{C.}
    Temporal clustering differences (on early versus late lists) versus feature
    clustering carryover. \textbf{All panels.} Each dot in the main
    scatterplots denotes the average scores for one participant. The colored
    regression lines are computed across participants. The inset displays
    condition-averaged results, where each dot reflects a single condition and
    the regression line is computed across experimental conditions. All error
    ribbons denote bootstrap-estimated 95\% confidence intervals.}
    \label{fig:clustering-carryover}
